---
title: "Data processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
# Load libraries that can't be/haven't been referenced using namespaces
library(tidyverse)
library(dplyr)
library(knitr)
library(ipumsr)
library(patchwork)

devtools::load_all("../../safpet")
```

# Overview 

In this vignette, we demonstrate how the two datasets, `data_df` and `data_to_fit_df` in `fit_model.Rmd` are preprocessed.

# Importing and processing the DHS data

Demographic and Health Survey (DHS) data that are used in this package are publicly available for free download after registering with the DHS Program at https://dhsprogram.com/data/. We import DHS data including all variables we need for our applications. Please see `fit_model.Rmd` for description of the variables. 

We first process the imported DHS data to produce `data_df` that consists of the entire population of interest regardless of TSLS. This dataset is used when calculating old and new measures for family planning.  

```{r}
ddi <- read_ipums_ddi("../data/idhs_00006.xml") 
dat <- read_ipums_micro(ddi)
dat <- dat %>% 
  mutate(FPMETHNOW_factor = as_factor(FPMETHNOW)) %>% 
  mutate(user = (FPMETHNOW_factor != "Not using")) %>%
  mutate(KIDDESIRE_factor = as_factor(KIDDESIRE)) %>% 
  mutate(inneed = (KIDDESIRE_factor == "Wants after 2+ years") | (KIDDESIRE_factor == "Wants no more")) %>%
  mutate(nosex = (AGE1STSEX == 0)) %>%
  #indicate those who had sex within the past two year: if TSLS less than 12 months, DHS records TSLS in days, weeks or months; max days reported is 97 days, max weeks reported is 52 weeks; hence, filter months < 24 months and year < 2 years 
  mutate(sa2yrs = (TIMESINCESEX < 324 | TIMESINCESEX == 401)) %>% 

  #subset to keep only women who do not want a child in the next 2 years, excluding Undecided/Wants, unsure timing/etc.
  filter(inneed) %>%
    filter(AGE >= 15 & AGE <= 49,
           MARSTAT == 10, #never married
           EXPOSURE1 == 0 #excludes pregnant, amenorrheic, infecund, menopausal women & missing & NIU
           ) %>%

  group_by(COUNTRY) %>%
    filter(YEAR == max(YEAR)) %>% #keep most recent surveys in each country
    ungroup() %>% 
  
  mutate(days = case_when(TIMESINCESEX >= 100 & TIMESINCESEX <= 197 ~ TIMESINCESEX - 100, #reported in days
                          TIMESINCESEX >= 201 & TIMESINCESEX <= 252 ~ (TIMESINCESEX - 200) * 7, #reported in weeks
                          TIMESINCESEX >= 301 & TIMESINCESEX <= 397 ~ (TIMESINCESEX - 300) * 30 + 1, #reported in months; this calculation matches with TIMESINCESEXD in the DHS data
                          TIMESINCESEX >= 401 & TIMESINCESEX <= 449 ~ (TIMESINCESEX - 400) * 365,
                          TRUE ~ NA_real_),   
  country = case_when(COUNTRY == 288 ~ "Ghana",
                      COUNTRY == 404 ~ "Kenya", 
                      COUNTRY == 566 ~ "Nigeria", 
                      COUNTRY == 716 ~ "Zimbabwe"
                      )) %>%
  
  filter(!is.na(country)) %>% 
  mutate(time = case_when(TIMESINCESEX >= 100 & 
                            TIMESINCESEX <= 197 ~ TIMESINCESEX - 100, 
                          TIMESINCESEX >= 201 & 
                            TIMESINCESEX <= 252 ~ TIMESINCESEX - 200,
                          TIMESINCESEX >= 301 & 
                            TIMESINCESEX <= 397 ~ TIMESINCESEX - 300,
                          TIMESINCESEX >= 401 & 
                            TIMESINCESEX <= 449 ~ TIMESINCESEX - 400,
                          TRUE ~ NA_real_),
         unitd = case_when(TIMESINCESEX >= 100 & TIMESINCESEX <= 197 ~ 1,
                          TIMESINCESEX >= 201 & TIMESINCESEX <= 252 ~ 7,
                          TIMESINCESEX >= 301 & TIMESINCESEX <= 397 ~ 31, 
                          TIMESINCESEX >= 401 & TIMESINCESEX <= 449 ~ 365,
                          TRUE ~ NA_real_),
         unit = factor(case_when(TIMESINCESEX >= 100 & 
                                   TIMESINCESEX <= 197 ~ "day",
                                TIMESINCESEX >= 201 & 
                                  TIMESINCESEX <= 252 ~ "week",
                                TIMESINCESEX >= 301 & 
                                  TIMESINCESEX <= 397 ~ "month",
                                TIMESINCESEX >= 401 & 
                                  TIMESINCESEX <= 449 ~ "year",
                          ), 
                       levels = c("day", "week", "month", "year"))) 

saveRDS(dat, file = here::here("data/dat_processed_paper.rds"))
```

We then select women whose TSLS < 2 years to produce `data_to_fit_df` which will be used for estimating the TBS density.

```{r}
dat_to_fit <- dat %>% 
  filter(TIMESINCESEX < 324 | TIMESINCESEX == 401) #selecting those who had sex within the past two year

saveRDS(dat_to_fit, file = here::here("data/dat_to_fit.rds"))
```





