---
title: "Model fitting"
output: html_document
date: "2022-1-5"
---

```{r setup, include=FALSE}
# Load libraries that can't be/haven't been referenced using namespaces
library(dplyr)
library(ggnewscale)
library(ggplot2)
library(haven)
library(rstan) 
library(splines2)
library(tidybayes)
library(tidyr)

options(mc.cores = parallel::detectCores())
rstan::rstan_options(auto_write = TRUE)

devtools::load_all("../../safpet")
```

# Overview 

In this vignette, we demonstrate how the functions in the R package are used to fit the model to the data set. After 
fitting the model, we compute the summary measures using the outputs (i.e. the posterior samples of model parameters).
Based on these summary measures, we quantify the frequencies of sex among users and non-users and use them to calculate 
the proposed new family planning measures (e.g. the unprotected exposure to risk of unintended pregnancy, the ratio of 
protected exposure over all women's exposure to risk of unintended pregnancy, etc). 

# The DHS data

We begin by summarizing the variables that we used to select the subsets of the DHS data for the population of interest.
The actual data pre-processing code/steps are available in `data_preprocessing.Rmd`. There are two datasets saved. 

- `data_df`: entire population regardless of TSLS. 
- `data_to_fit_df`: subset of `data_df` where TSLS < 2 years

`data_to_fit_df` is used for estimating the TBS density. the data set 
`data_df' is needed only for calculating the measures (we read it in when we use it, in the relevant section).

## Description of the variables

1. Primary outcome variable

- TIMESINCESEX (V527): TSLS data. We can get data $(z_i, u_i)$ from this variable as follows
\begin{align*}
   \text{ When } 100 \leq \text{ TIMESINCESEX } \leq 197 \text{ then } u[i] = 1 \text{ (reporting in days)} \text{ and } z_i = \text{ TIMESINCESEX } - 100,\\
   \text{ When } 201 \leq \text{ TIMESINCESEX } \leq 252 \text{ then } u[i] = 2 \text{ (reporting in weeks)} \text{ and } z_i = \text{ TIMESINCESEX } - 200,\\
      \text{ When } 301 \leq \text{ TIMESINCESEX } \leq 397 \text{ then } u[i] = 3 \text{ (reporting in months)} \text{ and } z_i = \text{ TIMESINCESEX } - 300,\\
         \text{ When } 401 \leq \text{ TIMESINCESEX } \leq 449 \text{ then } u[i] = 4 \text{ (reporting in years)} \text{ and } z_i = \text{ TIMESINCESEX } - 400\\
\end{align*}
- Note:
  - We select women who had sex within the past two years. 
  - Max days reported is 97 days, max weeks reported is 52 weeks 
  - We filter months <24 months and year <2 years for model fitting
  - If less than 12 months, DHS records in days, weeks or months
  
2. Other variables

- FPMETHNOW (V312): current method of FP. We indicate women who is "Not using" as non-users; all others are considered as users.
- AGE1STSEX (V525): age at first intercourse. We indicate women who have "Not had intercourse" (Code 00) as nosex women; later used to compute the proportion of no-sex.
- KIDDESIRE (V605): desire for more children. We include women who "Wants after 2+ years" or "Wants no more" (women in need), excluding Undecided/Wants, unsure timing/etc.
- EXPOSURE1 (V623): fecund, pregnant, amenorrheic, or infecund. We include women who are "Fecund", excluding pregnant, amenorrheic, infecund, menopausal women & missing & NIU.
- AGE (V012): age. We include women who are $\geq$ 15 and $\leq$ 49.
- MARSTAT (V501): woman's current marital or union status. We include women who are "Never married".
- COUNTRY: country. "Ghana","Kenya","Nigeria","Zimbabwe".
- YEAR: year of sample. Most recent survey in each country.
- PERWEIGHT (V005): sample weight for persons. Consider incorporating in the likelihood later?

## Visualization of the DHS data $(z_i, u_i)$

We can visualize the data as reported, $(z_i, u_i)$, in a histogram. We assign equal probability mass to the set of
$y_i$ associated with $z_i$. We use the encoding assumptions saved as object in `reporting_info_all.rds`. The following
shows an example of visualizing the reported data for Users in Ghana.

```{r}
data_to_fit_df <- readRDS(file = here::here("data/dat_to_fit.rds"))
data_to_fit_df
```

Check frequency.

```{r}
data_to_fit_df |>
  dplyr::group_by(unit, time) |>
  dplyr::tally() |>
  tidyr::spread(key = unit, value = n)
```

For plotting the data, we need to specify the day range and assumptions regarding reporting errors:

```{r}
specification <- generate_grids_and_units_specification()
```

```{r}
data_to_fit_df2 <- data_to_fit_df %>%
    mutate(
      reported_value = time,
      reported_unit = unit
    ) %>% 
  left_join(specification$df)
```

Plot data by reporting unit:
```{r}
ggout <- plot_fit(dat = data_to_fit_df2)
ggout
```

# The estimation procedure

## Model fitting 

Based on the data as reported, $(z_i, u_i)$ and the reporting issue assumptions (saved in specification), we fit the 
model. Below, we show an example of fitting the model for users in Ghana. 

```{r, results = "hide"}
# set seed to make results reproducible
set.seed(1234)

knots <- floor(c(seq(1, 14), seq(14 + 7, 6 * 7, 7), seq(6 * 7 + 30, max(specification$grid) - 30, 30)))

ghana_user_fit <- fit_density(
  data_to_fit_df,
  specification = specification, 
  ctry = "Ghana",
  use = TRUE,
  knots = knots,
  filename = "fits/ghana_user_fit", # if non-NULL, saved to file
  control = list(adapt_delta = 0.9)
)

ghana_non_user_fit <- fit_density(
  data_to_fit_df, 
  specification = specification, 
  ctry = "Ghana",
  use = FALSE,
  knots = knots,
  filename = "fits/ghana_nonuser_fit",# if non-NULL, saved to file
  control = list(adapt_delta = 0.9)
)
```

Plotting!

```{r}
ggout <- plot_fit(pmf_est = ghana_user_fit[["res"]], 
                  dat = ghana_user_fit[["dat"]],
                  title = "Users in Ghana")
ggout
```

```{r}
ggout <- plot_fit(pmf_est = ghana_non_user_fit[["res"]], 
                  dat = ghana_non_user_fit[["dat"]],
                  title="Non-users in Ghana")
ggout
```

```{r}
ggout <- plot_fit(pmf_est = ghana_non_user_fit[["res"]], 
                  dat = ghana_non_user_fit[["dat"]],
                  title="Non-users in Ghana")
ggout
```


## TBS samples from posterior samples 

We can obtain TBS samples from the posterior samples after fitting the model. We compute the summary measures of the 
estimated TBS distribution.

```{r}
# input: model fitting results for user
ghana_user_est_tbs <- estimate_tbs(ghana_user_fit$res) 
# input: model fitting results for non-user
ghana_non_user_est_tbs <- estimate_tbs(ghana_non_user_fit$res)

# input: samples of TBS means
tidybayes::median_qi(ghana_user_est_tbs$exptbs, .width = c(0.5, 0.8, 0.95))
tidybayes::median_qi(ghana_non_user_est_tbs$exptbs, .width = c(0.5, 0.8, 0.95))
```

## Old and new measures for family planning

Based on the summary measures of the estimated TBS distributions for users and non-users, we calculate the new measures for family planning. For each country, we also compute the existing (old) measures. 

For this we need the full data set: 
```{r}
data_df <- readRDS(file = here::here("data/dat_processed_paper.rds"))
data_df
```

```{r}
data_ghana_df <- data_df |>
  dplyr::filter(country == "Ghana")

 # input country data including both user and non-user
ghana_old_measures_df <- calculate_old_measures(data_ghana_df)
ghana_new_measures_df <- calculate_new_measures(data_ghana_df, ghana_non_user_fit$res, ghana_user_fit$res)
ghana_new_measures_ci_df <- calculate_new_measures(
  data_ghana_df, 
  ghana_non_user_fit$res, 
  ghana_user_fit$res, 
  ci_width = c(0.5, 0.8, 0.95)
)

measures_out_df <- tibble::tibble(ghana_old_measures_df, ghana_new_measures_df)
measures_add_df <- tibble::tibble(ghana_new_measures_ci_df)
measures_out_df
measures_add_df
```
